\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-09 chapters/abstract.tex
Comments François patients

Individuals with severe motor impairments, such as those with \ac{lis}
or eye-motor dysfunction, face substantial challenges when using traditional
gaze-dependent brain-computer interfaces \acp{bci}.
These systems require users to direct their gaze toward specific targets, a
task that becomes unfeasible for individuals with limited or no eye control.
This work addresses this limitation by developing gaze-independent
\ac{bci} methods, focusing on improving covert \ac{vsa}, where users can direct
their attention to a target without corresponding eye movements.
A key contribution of this work is the compensation for ERP latency jitter, a
variability that negatively impacts decoding performance in covert \ac{vsa}.
By handling this jitter, the proposed method enhances communication accuracy
without requiring gaze control, making \acp{bci} more usable for individuals with
motor impairments.

@@

The proposed methods were validated in experiments involving both healthy
individuals and 7 individuals with severe physical impairment and impaired eye-motor control.
Data were recorded from a control group, as well as from seven individuals with
gaze impairments.
These experiments demonstrated the robustness of the novel decoding methods,
showing that the system could effectively decode covert attention even when
direct gaze was impossible.

\paragraph{Keywords:} brain-computer interface, gaze independence, covert
visuospatial attention, event-related potentials, linear decoding, multilinear
decoding, motor impairment, gaze impairment

>>>>

Individuals with severe motor impairments, such as those with \ac{lis}
or eye-motor dysfunction, face substantial challenges when using traditional
gaze-dependent \acp{bci}.
These systems require users to direct their gaze toward specific targets, a
task that becomes unfeasible for individuals with limited or no eye control.
This work addresses this limitation by developing gaze-independent
\ac{bci} methods, focusing on improving covert \ac{vsa}, where users can direct
their attention to a target without corresponding eye movements.
A key contribution of this work is the compensation for ERP latency jitter, a
variability that negatively impacts decoding performance in covert \ac{vsa}.
By handling this jitter, the proposed method enhances communication accuracy
without requiring gaze control, making \acp{bci} more usable for individuals with
motor impairments.

@@

The proposed methods were validated in experiments involving both healthy
individuals and seven individuals with severe physical impairment and impaired eye-motor control.
These experiments demonstrated the robustness of the novel decoding methods,
showing that the system could effectively decode covert attention even when
direct gaze was impossible.

\paragraph{Keywords:} \acl{bci}, gaze independence, covert
visuospatial attention, event-related potentials, linear decoding, multilinear
decoding, motor impairment, gaze impairment

### 2024-10-09 chapters/abstract_fr.tex
Comments François patients

\chapter*{R\'esum\'e}
Les individus ayant des handicaps moteurs sévères, tels que ceux atteints de \ac{lis} ou de dysfonctionnement oculomoteur, rencontrent des défis considérables lors de l'utilisation d'interfaces cerveau-ordinateur \acp{bci} traditionnelles.
Ces systèmes nécessitent que les utilisateurs dirigent leur regard vers des cibles spécifiques, une tâche qui devient impraticable pour les personnes ayant un contrôle oculaire limité ou inexistant.
Ce travail aborde cette limitation en développant des méthodes de \ac{bci} indépendantes du regard, en se concentrant sur l'amélioration de la \ac{vsa} cachée, où les utilisateurs peuvent diriger leur attention vers une cible sans mouvements oculaires correspondants.
Une contribution clé de ce travail est la compensation pour le jitter de latence ERP, une variabilité qui impacte négativement la performance de décodage dans la \ac{vsa} cachée.
En gérant ce jitter, la méthode proposée améliore la précision de la communication sans nécessiter de contrôle du regard, rendant les \acp{bci} plus utilisables pour les personnes ayant des handicaps moteurs.

Au-delà de l'indépendance du regard, le travail fait également progresser les techniques générales de décodage \ac{erp} en affinant la structure des estimateurs linéaires et multilinaires.
Ces méthodes améliorent la précision de classification dans une gamme de conditions \ac{bci}, en particulier lorsque les données d'entraînement sont limitées.
L'introduction de régularisation structurée dans les modèles linéaires et multilinaires améliore l'interprétabilité des classificateurs et réduit le temps d'entraînement ainsi que la complexité computationnelle.
Cela permet un entraînement plus efficace et une meilleure performance sur des données non vues, contribuant à des systèmes plus fiables qui s'adaptent à divers contextes et besoins des utilisateurs.

Les méthodes proposées ont été validées lors d'expériences impliquant à la fois des individus en bonne santé et 7 individus avec des handicaps physiques sévères et un contrôle oculomoteur altéré.
Les données ont été enregistrées à partir d'un groupe de contrôle, ainsi que de sept individus ayant des impairments du regard.
Ces expériences ont démontré la robustesse des nouvelles méthodes de décodage, montrant que le système pouvait efficacement décoder l'attention cachée même lorsque le regard direct était impossible.

\paragraph{Mots-clés :} interface cerveau-ordinateur, indépendance du regard, attention visuo-spatiale cachée, potentiels évoqués liés aux événements, décodage linéaire, décodage multilinéraire, handicap moteur, handicap du regard.

>>>>

\chapter*{R\'esum\'e}
Les individus ayant des handicaps moteurs sévères, tels que ceux atteints de
syndrome d'enfermement ou de dysfonctionnement oculomoteur, rencontrent des défis considérables
lors de l'utilisation d'interfaces cerveau-ordinateur (BCI)  traditionnelles.
Ces systèmes nécessitent que les utilisateurs dirigent leur regard vers des
cibles spécifiques, une tâche qui devient impraticable pour les personnes ayant un contrôle oculaire limité ou inexistant.
Ce travail vise \`a palier cette limitation en développant des méthodes de BCI
indépendantes du regard, en se concentrant sur l'amélioration de la attention visuo-spatiale (VSA)
cachée,
où les utilisateurs peuvent diriger leur attention vers une cible sans mouvements
oculaires correspondants.
Une contribution clé de ce travail est la compensation pour la compensation de
la variabilité de latence des potentiels évoqués (ERP), laquelle impacte
la performance de décodage dans la
VSA cachée.
En gérant cette variabilite, la méthode proposée améliore la précision de la communication
sans nécessiter de contrôle du regard, rendant les BCI plus utilisables pour les personnes ayant des handicaps moteurs.

Au-delà de l'indépendance du regard, le travail fait également progresser les
techniques générales de décodage ERP en affinant la structure des estimateurs
linéaires et multilinéaires.
Ces méthodes améliorent la précision de classification dans une gamme de
conditions BCI, en particulier lorsque les données d'entraînement sont limitées.
L'introduction de régularisation structurée dans les modèles linéaires et
multilinéaires améliore l'interprétabilité des classificateurs et réduit le temps d'entraînement ainsi que la complexité computationnelle.
Cela permet un entraînement plus efficace et une meilleure performance sur des données non vues, contribuant à des systèmes plus fiables qui s'adaptent à divers contextes et besoins des utilisateurs.

Les méthodes proposées ont été validées lors d'expériences impliquant à la fois
des individus en bonne santé et sept individus avec des handicaps physiques sévères et un contrôle oculomoteur altéré.
Ces expériences ont démontré la robustesse des nouvelles méthodes de décodage, montrant que le système pouvait efficacement décoder l'attention cachée même lorsque l'observation directe de la cible  était impossible.

\paragraph{Mots-clés :} interface cerveau-ordinateur, indépendance du regard, attention visuo-spatiale cachée, potentiels évoqués liés aux événements, décodage linéaire, décodage multilinéaire, handicap moteur, handicap du regard.

### 2024-10-09 chapters/bci.tex
Comments François patients

\chapter{Brain-computer interfaces}
\label{sec:bci}
\todo{search for ??}
\todofc{double check statements about what has been published 'for the sake of clarity, the introduction has been mofed to"}
\todo{fix all latex warnings to make sure no refs are missing}
\todo{uniform ref or cref}
\todo{consistently terminology individuals with severe
speech and physical impairments (SSPI), avoid phrasing 'suffering from'}
\todohsm{more subtitles}
\todo{uniform vsa condition vsa setting}
\todo{are all acronyms defined/Acro corectly used in chapters?}
\todo{acronyms written in full every new chapter and in captions?}
\todo{are textcites used when necessary?}
\todo{fix lacheck/chktex notifications}
\todo{footnot skips a page in wcble chap}

@@

The key problem for severely paralyzed patients is that the mind wants to go where the
body cannot follow.
Couldn't we, therefore, design a system that directly interacts with the mind?
This system can then form an interface between the mind and any kind of technology, like
a robot arm or speech synthesizer, interacting with the world on the user's behalf.
Such a device is the \ac{bci}\footnote{Sometimes also termed brain-machine interface
(BMI) when coupled to a physical actuator, like a robot arm or an exoskeleton.
The term \ac{bci} is usually preferred for assistive technology and communication
devices.}, a system that directly couples actions in the `real world' to the mental state
of the user.

Starting in the 1970s from methods to establish basic communication from minimal brain
activity (yes/no, left/right, \ldots)~\cite{Wolpaw2002}, \acp{bci} have gradually evolved
into sophisticated interaction schemes that restore or enhance many capabilities by
bypassing defective parts of the nervous or muscle system.
By now, classic \acp{bci} methods are starting to mature outside the lab setting and can
be used as assistive communication technology by patients with impaired
speech~\cite{Wolpaw2018,Milekovic2018}.
Impressive cutting-edge experimental examples include a brain-spine interface allowing a
paraplegic patient to walk again~\cite{Lorach2023}, a \ac{bci} translating internal
speech to a facial avatar speaking the imagined text~\cite{Metzger2023}, and fast
communication through decoding imagined handwritten symbols~\cite{Willett2021}.
Recently, the idea has also gained popular traction through Elon Musk's
NeuraLink~\cite{Musk2019} initiative, which imagines a multipurpose \ac{bci}.

In general, devices processing direct inputs from the central nervous system are useful
for rehabilitation and medical diagnosis or treatment.
Additionally, they bring a new paradigm to human-computer interaction, especially when
paired with virtual or augmented reality~\cite{SiMohammed2017}.
Ultimately, they have always been especially promising as assistive technology for
paralyzed patients limited in their communication ability.
It is this \ac{bci}-controlled communication where our focus lies, as it offers a
transformative means for individuals with severe motor impairments to regain functional
independence to write and speak through assistive technology.
This allows them to talk to their loved ones and caregivers, exercise hobbies, and makes
many aspects of daily life more accessible.

@@

The brain's activity can be recorded using various neuroimaging technologies.
These can range from brain scans~\cite{Weiskopf2004} using functional magnetic resonance
imaging (fMRI) to more portable technologies like acoustic signals obtained by
functional ultrasound imaging (fUS)~\cite{Zheng2023} and near-infrared spectroscopy
(fNIRS)~\cite{Borgheai2020}.

@@

The \emph{invasiveness} of the technology forms an important concept in determining its
suitability for a given application and patient.
As illustrated in \cref{fig:bci/recording}, invasive electrodes can often measure a
very specific brain region and can result in better signal quality, at the cost of the
risks involved with brain surgery.

\fullpagefig[topleft]{%
  \includegraphics[width=\fullpagefigwidth]{figures/bci/recording_modalities.pdf}
}{%
  Various electrophysiology recording technologies.
}{%
	Various electrophysiology recording technologies and their respective position in a
	cross-section of the skull.
	There is a trade-off between invasiveness and spatial resolution.
}{%
  fig:bci/recording
}

@@

Larger implanted electrodes are referred to as intracranial \ac{eeg} (i\ac{eeg}).
Depth electrodes such as those used in stereo-\ac{eeg} \ac{bci}~\cite{Wu2024} tasks and
closed-loop adaptive control for Deep Brain Stimulation to mitigate Parkinson's disease
symptoms~\cite{Arlotti2018}.
Depth electrodes are longer rods with multiple electrodes that can extend deeper into
the brain.
Since these still penetrate the cortex, electrodes that sit on top of the cortex are
more popular, as they are considered less invasive.
\Ac{ecog} is a powerful method with many applications in \ac{bci}~\cite{Schalk2011} and
epilepsy diagnosis.
Newer developments like micro-ECoG (µECoG)~\cite{Shokoueinejad2019}, with a higher
number of smaller electrodes per surface area, allow for more precise measurements.

@@

\Ac{eeg} is the de facto non-invasive electrophysiology standard in clinical neurology
practice as well as in \ac{bci} research.
Developed since 1924~\cite{Berger1929}, it is cheap and relatively practical and offers
great temporal resolution up to thousands of Hz.
\Ac{eeg} measures the electrical potential over time at a given electrode relative to a
reference electrode in volts (V), often on the scale of microvolts ($\mu$V).
As it is easily applicable on the outside of the scalp, many practical consumer-grade
\ac{eeg} headsets exist in the form of (often wireless) electrode caps or helmets.
Current consumer systems sometimes feature dry electrodes, but clinical and research
systems often use active electrodes with preamplifiers and electrolyte gel to reduce
impedance for improved signal quality, which slightly decreases practicality, since they
require a trained individual to properly administer.
Nevertheless, it is the most accessible \ac{bci} technology for both patients and
researchers.

@@

Despite its noisy nature, \ac{eeg} is the recording methodology of choice for our
\ac{bci} because of its wide acceptance by patients.
To counteract the noise present in our recording, we must on the one hand evoke strong,
informative brain activity using a suited \ac{bci} paradigm, and on the other hand pick
a suited decoder which can isolate the signal of interest and filter out the noise.

@@

\fullpagefig{%
  \input{figures/bci/paradigms.tikz.tex}
}{%
  Overview of \acs{bci} paradigms.
}{%
  A \ac{bci} paradigm defines which brain activity will be used for control.
  They can be categorized in active, reactive and passive paradigms, depending on their
  reliance on external stimulation and active participation required from the user.
  Adapted from~\textcite{Muehl2014}.
}{%
  fig:bci/paradigms
}

@@

Decoder choice is heavily dependent on constraints set by the paradigm and the
application.
The number of output degrees of features, the serial or parallel nature of performed
tasks, and whether the paradigm is active or reactive, all need to be taken into
account.
One of these choices is whether to solve a \emph{regression} or \emph{classification}
problem.
In regression, a continuous output variable is predicted for a new sample, while
classification attempts to sort it into one of a set of discrete classes.
Regression can be useful in \acp{bci} for applications like mapping imagined movement to
an external robot arm.
For communication \acp{bci}, however, conveying information in a symbolic manner is
often of interest.
Many applications resemble virtual keyboards or map a user's actions to discrete words,
predefined actions, or letters for full control.
Hence, such problems often present themselves as classification tasks.

@@

The visual oddball paradigm is a \emph{reactive}, \emph{stimulus-dependent},
\ac{bci} paradigm with all the benefits and drawbacks.
Nevertheless, it can score relatively high on the \ac{itr} parameters established above.
The brain response of interest is the visual \ac{erp}, which can be accurately measured
and decoded from non-invasive \ac{eeg} signals with a relatively low computational
effort and short calibration time.
First established by~\cite{Farwell1988} in 1988, it is well supported by literature and
has been proven to work for patients in day-to-day home use~\cite{Wolpaw2018}.

@@

  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/bci/erp.tikz.tex}
    \caption{The idealized \ac{erp} evoked by an attended stimulation in a single
    channel.}
  \end{subfigure}%
\end{figure}

@@

These decoders define the last building blocks of our visual oddball \ac{bci}.
Yet, the overview until now has been rather theoretical.
Interesting and complex challenges arise when applying these concepts outside of the lab
in augmented and assistive communication technologies for patients with \acp{sspi}.

@@

This leads to a problem that is not often addressed:
a visual \ac{bci} is dependent on the gaze of the user, and hence on muscle control.
This is exactly what \acp{bci} for augmented and assistive technology try to avoid.
The \ac{bci} target population consists of individuals who suffer from various degrees
of paralysis, or even from \ac{lis}, the complete loss of muscle control with preserved
consciousness.
While \ac{bci}s are most attractive as a solution for individuals with \ac{sspi},
studies paradoxically often report poor performance in patient populations precisely due
to the lack of adequate eye motor control.
Equally often, participants with gaze impairments were excluded from the study.

Different degrees of eye motor impairment can render visual \ac{bci}s uncomfortable or
outright impossible to use.
When operating a visual \ac{bci}, the usual rapid series of forced saccades followed by
fixation is tiring over time, even for healthy control subjects.
A suitable alternative would allow the user to keep their eyes in an at-rest position of
their choice while operating the \ac{bci}.
This points to the need for gaze-independent solutions, which we will explore further in
\cref{sec:gaze-independence} and the rest of this thesis.

>>>>

\chapter{Brain-computer interfaces}
\label{sec:bci}
\todo{consistently terminology individuals with severe
speech and physical impairments (SSPI)}
\todo{uniform vsa condition vsa setting}
\todo{are all acronyms defined/Acro corectly used in chapters?}
\todo{footnot skips a page in wcble chap}

@@

The key problem for severely paralyzed individuals is that the mind wants to go where the
body cannot follow.
Couldn't we, therefore, design a system that directly interacts with the mind?
This system can then form an interface between the mind and any kind of technology, like
a robot arm or speech synthesizer, interacting with the world on the user's behalf.
Such a device is the \ac{bci}\footnote{Sometimes also termed brain-machine interface
(BMI) when coupled to a physical actuator, like a robot arm or an exoskeleton.
The term \ac{bci} is usually preferred for assistive technology and communication
devices.}, a system that directly couples actions in the `real world' to the mental state
of the user.

Starting in the 1970s from methods to establish basic communication from minimal brain
activity (yes/no, left/right, \ldots)~\cite{Wolpaw2002}, \acp{bci} have gradually evolved
into sophisticated interaction schemes that restore or enhance many capabilities by
bypassing defective parts of the nervous or muscle system.
By now, classic \acp{bci} methods are starting to mature outside the lab setting and can
be used as assistive communication technology by individuals with \acl{sspi}.
speech~\cite{Wolpaw2018,Milekovic2018}.
Impressive cutting-edge experimental examples include a brain-spine interface allowing a
paraplegic individual to walk again~\cite{Lorach2023}, a \ac{bci} translating internal
speech to a facial avatar speaking the imagined text~\cite{Metzger2023}, and fast
communication through decoding imagined handwritten symbols~\cite{Willett2021}.
Recently, the idea has also gained popular traction through Elon Musk's
NeuraLink~\cite{Musk2019} initiative, which imagines a multipurpose \ac{bci}.

In general, devices processing direct inputs from the central nervous system are useful
for rehabilitation and medical diagnosis or treatment.
Additionally, they bring a new paradigm to human-computer interaction, especially when
paired with virtual or augmented reality~\cite{SiMohammed2017}.
Ultimately, they have always been especially promising as assistive technology for
people limited in their communication ability.
It is this \ac{bci}-controlled communication where our focus lies, as it offers a
transformative means for individuals with severe motor impairments to regain functional
independence to write and speak through assistive technology.
This allows them to talk to their loved ones and caregivers, exercise hobbies, and makes
many aspects of daily life more accessible.

@@

The brain's activity can be recorded using various neuroimaging technologies.
These can range from brain scans~\cite{Weiskopf2004} using functional magnetic resonance
imaging (fMRI) to more portable technologies like acoustic signals obtained by
functional ultrasound imaging (fUS)~\cite{Zheng2023} and functional near-infrared spectroscopy
(fNIRS)~\cite{Borgheai2020}.

@@

The \emph{invasiveness} of the technology forms an important concept in determining its
suitability for a given application and user.
As illustrated in \cref{fig:bci/recording}, invasive electrodes can often measure a
very specific brain region and can result in better signal quality, at the cost of the
risks involved with brain surgery.

\fullpagefig[topleft]{%
  \includegraphics[width=\fullpagefigwidth]{figures/bci/recording_modalities.pdf}
}{%
  Various electrophysiology recording technologies.
}{%
	Various electrophysiology recording technologies and their respective position in a
	cross-section of the skull.
	There is a trade-off between invasiveness and spatial resolution.
}{fig:bci/recording}

@@

Larger implanted electrodes are referred to as intracranial \ac{eeg} (i\ac{eeg}).
Depth electrodes, like those found in in stereo-\ac{eeg}, can be used for \ac{bci}~\cite{Wu2024}
 and other applications, like closed-loop adaptive control for Deep Brain Stimulation to mitigate Parkinson's disease
symptoms~\cite{Arlotti2018}.
Depth electrodes are longer rods with multiple electrodes that can extend deeper into
the brain.
Since these still penetrate the cortex, electrodes that sit on top of the cortex are
more popular, as they are considered less invasive.
\Ac{ecog} is a powerful method with many applications in \ac{bci}~\cite{Schalk2011} and
epilepsy diagnosis.
Newer developments like micro-ECoG (µECoG)~\cite{Shokoueinejad2019}, with a higher
number of smaller electrodes per surface area, allow for more precise measurements.

@@

\Ac{eeg} is the de facto non-invasive electrophysiology standard in clinical neurology
practice as well as in \ac{bci} research.
Developed since 1924~\cite{Berger1929}, it is cheap and relatively practical and offers
great temporal sampling resolution up to thousands of Hz.
\Ac{eeg} measures the electrical potential over time at a given electrode relative to a
reference electrode in volts (V), often on the scale of microvolts ($\mu$V).
As it is easily applicable on the outside of the scalp, many practical consumer-grade
\ac{eeg} headsets exist in the form of (often wireless) electrode caps or helmets.
Current consumer systems sometimes feature dry electrodes, but clinical and research
systems often use active electrodes with preamplifiers and electrolyte gel to reduce
impedance for improved signal quality, which slightly decreases practicality, since they
require a trained individual to properly administer.
Nevertheless, it is the most accessible \ac{bci} technology for both users and
researchers.

@@

Despite its noisy nature, \ac{eeg} is the recording methodology of choice for our
\ac{bci} because of its wide acceptance by users.
To counteract the noise present in our recording, we must on the one hand evoke strong,
informative brain activity using a suited \ac{bci} paradigm, and on the other hand pick
a suited decoder which can isolate the signal of interest and filter out the noise.

@@

\fullpagefig{%
  \input{figures/bci/paradigms.tikz.tex}
}{%
  Overview of \acs{bci} paradigms.
}{%
  A \ac{bci} paradigm defines which brain activity will be used for control.
  They can be categorized in active, reactive and passive paradigms, depending on their
  reliance on external stimulation and active participation required from the user.
  Adapted from~\textcite{Muehl2014}.
}{fig:bci/paradigms}

@@

Decoder choice is heavily dependent on constraints set by the paradigm and the
application.
The number of output degrees of freedom, the serial or parallel nature of performed
tasks, and whether the paradigm is active or reactive, all need to be taken into
account.
One of these choices is whether to solve a \emph{regression} or \emph{classification}
problem.
In regression, a continuous output variable is predicted for a new sample, while
classification attempts to sort it into one of a set of discrete classes.
Regression can be useful in \acp{bci} for applications like mapping imagined movement to
an external robot arm.
For communication \acp{bci}, however, conveying information in a symbolic manner is
often of interest.
Many applications resemble virtual keyboards or map a user's actions to discrete words,
predefined actions, or letters for full control.
Hence, such problems often present themselves as classification tasks.

@@

The visual oddball paradigm is a \emph{reactive}, \emph{stimulus-dependent},
\ac{bci} paradigm with all the benefits and drawbacks.
Nevertheless, it can score relatively high on the \ac{itr} parameters established above.
The brain response of interest is the visual \ac{erp}, which can be accurately measured
and decoded from non-invasive \ac{eeg} signals with a relatively low computational
effort and short calibration time.
First established by~\cite{Farwell1988} in 1988, it is well supported by literature and
has been proven to work for those with \acl{sspi} in day-to-day home use~\cite{Wolpaw2018}.

@@

  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/bci/erp.tikz.tex}
    \caption{The idealized \ac{erp} evoked by an attended stimulation in a single
    channel.}
    \label{fig:bci/erp}
  \end{subfigure}%
\end{figure}

@@

These decoders define the last building blocks of our visual oddball \ac{bci}.
Yet, the overview until now has been rather theoretical.
Interesting and complex challenges arise when applying these concepts outside of the lab
in augmented and assistive communication technologies for users with \acl{sspi}.

@@

This leads to a problem that is not often addressed:
a visual \ac{bci} is dependent on the gaze of the user, and hence on muscle control.
This is exactly what \acp{bci} for augmented and assistive technology try to avoid.
The \ac{bci} target population consists of individuals with severe physical
impairment due to  with various degrees
of paralysis, or even with \ac{lis}, the complete loss of muscle control with preserved
consciousness.
While \ac{bci}s are most attractive as a solution for individuals with \ac{sspi},
studies paradoxically often report poor performance in exactly this group,
precisely due to their lack of adequate eye motor control.
Equally often, participants with gaze impairments were excluded from the study.

Different degrees of eye motor impairment can render visual \ac{bci}s uncomfortable or
outright impossible to use.
When operating a visual \ac{bci}, the usual rapid series of forced saccades followed by
fixation is tiring over time, even for healthy control subjects.
A suitable alternative would allow the user to keep their eyes in an at-rest position of
their choice while operating the \ac{bci}.
This points to the need for gaze-independent solutions, which we will explore further in
\cref{sec:gaze-independent} and the rest of this thesis.

### 2024-10-09 chapters/bttda.tex
Comments François patients

\chapter{Block-term tensor discriminant analysis}
\label{sec:bttda}
\input{chapters/bttda/tensorviz.tex}
\emph{This chapter was submitted as~\textcite{VanDenKerchove2024a}}.

@@

\subsection{Block contribution}
To analyze the contribution of extra feature blocks extracted by BTTDA over
the first one found by HODA, we pick an ERP ($K=2$) dataset
(BNCI2014-008) and an MI ($K=3$) dataset (BNCI2014-001).
We report within-session ROC-AUC scores for training, validation, and test data as a function
of the number of blocks increases, shown in \cref{fig:blocks}.
Training and validation folds were taken from the model selection procedure.
Additionally, the Normalized Mean Squared Error (NMSE) is reported for the
reconstructed from the truncated BTTDA decomposition
$\textstyle{\hat{\ten{X}}^{(B)}=\sum_b^B\ten{G}^{(b)}\mmpr{\mat{U}^{(b)}}}$.
NMSE is calculated as:
\begin{equation}
  \nmse\left(\ten{X}, \hat{\ten{X}}^{(B)}\right) =
	\frac{\sum_n^N\left\lVert\ten{X}(n)-\ten{X}_\text{rec}^{(B)}(n)\right\rVert_F^2}
	{\sum_n^N\left\lVert\ten{X}(n)\right\rVert_F^2}
\end{equation}
\begin{figure*}[t]
	\input{figures/bttda/blocks.pgf}
  \caption[Analysis of NMSE and classification score per block.]{%
    Normalized Mean Square Error (NMSE) (left), and difference in
    classification score for the training and validation for the ERP datasets
    (top row) and the MI dataset (bottom row)	of the greedy model selection
    procedure (right), as a function of the number of BTTDA blocks $b$.
		While NMSE monotonically decreases for the evaluated datasets, better class
		separation will be achieved, but eventually overfitting occurs and validation
		and test scores will drop, or plateau due to feature selection.
	}
	\label{fig:blocks}
\end{figure*}
\todo{format figure}

@@

Due to its heuristic nature, the greedy algorithm does not always result in the
set of ranks with the highest achievable performance.
In combination with the fact that the feature selection cutoff parameter was
fixed somewhat arbitrarily, it is clear that thorough
hyperparameter optimization could improve performance.
Furthermore, it is clear that our proposed model selection procedure does not
necessarily result in an optimal set of blocks that group coherent projections
within the same block, according to some desirable metric.
Examples of this are sparsity, pattern interpretability, minimal or maximal
between- or within-block feature correlation, decreasing discriminability, etc.
Finally, the presented results are limited by our artificial restriction
that $(r_1=r_2=\ldots=r_k)$ to reduce the computational demand of the presented
experiments.
In a sense, this goes against the proposition of increased model flexibility.
It might well be that BTTDA in some cases offers little added value over the
Tucker-structured HODA, when both are given free choice of rank.
Future efforts should focus on automatic parameter setting, e.g., using sparsity
or information criteria such as the ones used in BTTR~\cite{Faes2022} or other
statistical measures based on the model's application.

>>>>

\chapter{Block-term tensor discriminant analysis}
\label{sec:bttda}
\input{chapters/bttda/tensorviz.tex}
\emph{This chapter was submitted as~\textcite{VanDenKerchove2024a}.}

@@

\subsection{Block contribution}
To analyze the contribution of extra feature blocks extracted by BTTDA over
the first one found by HODA, we pick an ERP ($K=2$) dataset
(BNCI2014-008) and an MI ($K=3$) dataset (BNCI2014-001).
We report within-session ROC-AUC scores for training, validation, and test data as a function
of the number of blocks increases, shown in \cref{fig:blocks}.
Training and validation folds were taken from the model selection procedure.
Additionally, the Normalized Mean Squared Error (NMSE) is reported for the
reconstructed from the truncated BTTDA decomposition
$\textstyle{\hat{\ten{X}}^{(B)}=\sum_b^B\ten{G}^{(b)}\mmpr{\mat{U}^{(b)}}}$.
NMSE is calculated as:
\begin{equation}
  \nmse\left(\ten{X}, \hat{\ten{X}}^{(B)}\right) =
	\frac{\sum_n^N\left\lVert\ten{X}(n)-\ten{X}_\text{rec}^{(B)}(n)\right\rVert_F^2}
	{\sum_n^N\left\lVert\ten{X}(n)\right\rVert_F^2}
\end{equation}
\begin{figure*}[t]
  \centering
  \makebox[\linewidth][c]{
	  \input{figures/bttda/blocks.pgf}
  }
  \caption[Analysis of NMSE and classification score per block.]{%
    Normalized Mean Square Error (NMSE) (left), and difference in
    classification score for the training and validation for the ERP datasets
    (top row) and the MI dataset (bottom row)	of the greedy model selection
    procedure (right), as a function of the number of BTTDA blocks $b$.
		While NMSE monotonically decreases for the evaluated datasets, better class
		separation will be achieved, but eventually overfitting occurs and validation
		and test scores will drop, or plateau due to feature selection.
	}
	\label{fig:blocks}
\end{figure*}
\todo{format figure}

@@

Due to its heuristic nature, the greedy algorithm does not always result in the
set of ranks with the highest achievable performance.
In combination with the fact that the feature selection cutoff parameter was
fixed somewhat arbitrarily, it is clear that thorough
hyperparameter optimization could improve performance.
Furthermore, it is clear that our proposed model selection procedure does not
necessarily result in an optimal set of blocks that group coherent projections
within the same block, according to some desirable metric.
Examples of this are sparsity, pattern interpretability, minimal or maximal
between- or within-block feature correlation, decreasing discriminability, etc.
Finally, the completeness of the presented results is limited by the artificial
restriction in hyperparameter choice.
We imposed $(r_1=r_2=\ldots=r_k)$ that to reduce the computational demand of
the performed experiments.
In a sense, this goes against the proposition of increased model flexibility.
It might well be that BTTDA in some cases offers little added value over the
Tucker-structured HODA, when both are given free choice of rank.
Future efforts should focus on automatic parameter setting, e.g., using sparsity
or information criteria such as the ones used in BTTR~\cite{Faes2022} or other
statistical measures based on the model's application.

### 2024-10-09 chapters/conclusion.tex
Comments François patients

Together, these results show that there is an interest in developing a new class of ERP-BCI
interfaces for patients that prefer to choose their own gaze strategy, to
avoid the effort of redirecting their eye gaze to different spatial locations
on the stimulation screen in manners that might be uncomfortable for them.
Coincidentally, advances in gaze-independent decoding also build towards a
solution for the Midas Touch Problem in visual \ac{bci}.
Here, a \ac{bci} user sometimes accidentally selects a target while not intending
to give any input.
Decoding of true intention independent of eye
gaze, with the option of gazing without paying visuospatial attention to a
stimulus, would counteract this.

@@

Finally, we wish to expand our experimentation with individuals with
\acp{sspgi}.
We aim to revisit the patients with an on-line experiment and comfort
assessment to establish the usability of the proposed interface and compare its
\ac{itr} to the state-of-the-art literature.

@@

The optimal approach should unify decoder development, paradigm choice, and interface
design, as advocated by~\textcite{Pan2022} and~\textcite{Fouad2020}, by integrating both
engineering and clinical perspectives.
Performance and abilities likely depend to a great degree on the user,
so off-line decoding performance in healthy controls is probably not a good
predictor of on-line decoding performance and \ac{itr} in individuals with
\ac{sspi}.
Incremental gains in decoding performance for a specific paradigm might
also be outweighed by how well that paradigm is adapted to the user, factoring in
their skills and abilities.
Longitudinal research should follow several patient groups, specifically those with gaze
impairment, and experiment with specific paradigms, interfaces, and decoders for
a given \ac{bci} user.

>>>>

Together, these results show that there is an interest in developing a new class of ERP-BCI
interfaces for users that prefer to choose their own gaze strategy, to
avoid the effort of redirecting their eye gaze to different spatial locations
on the stimulation screen in manners that might be uncomfortable for them.
Coincidentally, advances in gaze-independent decoding also build towards a
solution for the Midas Touch Problem in visual \ac{bci}.
Here, a \ac{bci} user sometimes accidentally selects a target while not intending
to give any input.
Decoding of true intention independent of eye
gaze, with the option of gazing without paying visuospatial attention to a
stimulus, would counteract this.

@@

Finally, we wish to expand our experimentation with individuals with
\acp{sspgi}.
We aim to revisit the participants with an on-line experiment and comfort
assessment to establish the usability of the proposed interface and compare its
\ac{itr} to the state-of-the-art literature.

@@

The optimal approach should unify decoder development, paradigm choice, and interface
design, as advocated by~\textcite{Pan2022} and~\textcite{Fouad2020}, by integrating both
engineering and clinical perspectives.
Performance and abilities likely depend to a great degree on the user,
so off-line decoding performance in healthy controls is probably not a good
predictor of on-line decoding performance and \ac{itr} in individuals with
\ac{sspi}.
Incremental gains in decoding performance for a specific paradigm might
also be outweighed by how well that paradigm is adapted to the user, factoring in
their skills and abilities.
Longitudinal research should follow several user groups, specifically those with gaze
impairment, and experiment with specific paradigms, interfaces, and decoders for
a given \ac{bci} user.

### 2024-10-09 chapters/conflict.tex
Comments François patients

\chapter*{Conflict of interest}
The authors declare no conflict of interest.

>>>>

\chapter*{Conflict of interest \& funding}
The authors declare no conflict of interest.

Arne Van Den Kerchove is funded by a bilateral cooperation program between KU
Leuven (special research fund of the KU Leuven GPUDL/20/031) and the university
of Lille under the Global PhD Partnership program.
Prof. Marc Van Hulle is supported by research grants received from
the European Union’s Horizon Europe Marie Sklodowska-Curie Action program
(grant agreement No. 101118964), the European Union’s Horizon 2020 research and
innovation program (grant agreement No. 857375), the special research fund of
the KU Leuven (C24/18/098), the Belgian Fund for Scientific Research – Flanders
(G0A4118N, G0A4321N, G0C1522N), and the Hercules
Foundation (AKUL 043).
